Transparency of COVID-19-related research in medical journals

### 2022-04-27 manuscript/manuscript_covid_transparency.qmd
First draft

format: html
#format:
#  pdf:
#    colorlinks: true
editor: visual
bibliography: references.bib
csl: vancouver.csl
---

@@

**Results:** The total number of COVID-19-related articles was 175 550, of which we were able to retrieve full texts of 132 618 (75.6%) articles from the database Of those, 87.0%, (95% CI: 86.8%--87.2%) provided conflicts of interest disclosure and 68.9% (95% CI: 68.7%--69.2%) funding disclosure and 3.6%, (95% CI: 3.5%--3.7%) had registered protocol. 8.7% (95% CI: 8.6%--8.9%) of studies shared raw data and 3.1% (95% CI: 3.0%--3.2%) of studies shared code. Transparent practices were more common in articles published in 2021 than in 2020.

@@

During the recent COVID-19 pandemic, part of the public in developed countries justified their lack of adherence to public health measures due to the lack of transparency of the scientific studies used as an argument by national authorities [@Skipper2020-rs\]. Thus, supranational bodies such as the OECD indicate that transparent communication to citizens is vital to support public health measures and to counter misinformation \[@Bellantoni_undated-tf\]. On the other hand, the COVID-19 pandemic launched an incredibly fast and voluminous avalanche of scientific publications across the different fields of science. This research led to the development of vaccines as well as the evaluation of community interventions. On the one hand, it is possible that transparent scientific practices, like data and code sharing, have helped to fight pandemic globally \[@besançon2021; @vuong2022\], while on the other hand, the lack of transparency of some key developments has led the public to mistrust the results of research and public health measures \[@Skipper2020-rs\]. This shows that there is a lack of awareness about the level of transparency practices in COVID-19-related medical literature. As an example, it is unknown whether open science initiatives related to COVID-19 pandemic \[@besançon2021; @vuong2022] to advance data sharing, has led to higher number of articles sharing data in COVID-19 research compared to what has been seen in studies in general in biomedical literature [@serghiou2021].

@@

We searched for open access COVID-19-related PubMed-indexed articles available in the EPMC database. We restricted our search to papers in English and considered papers published from 2020-01-01 to 2021-12-31 with variants of COVID-19 keyword in their title, keywords, or results. This enabled us to bypass screening for COVID-19-related papers and decrease the human labor for our research. We confirmed our approach by choosing a random sample of 100 papers and of those, all were relevant to COVID-19.

@@

(COVID-19 query for title, keywords, and results) AND (SRC:"MED") AND (LANG:"eng" OR LANG:"en" OR LANG:"us") AND (FIRST_PDATE:\[2020-01-01 TO 2021-12-31\]) AND (OPEN_ACCESS:y) AND (PUB_TYPE:"Journal Article" OR PUB_TYPE:"research-article" OR PUB_TYPE:"rapid-communication" OR PUB_TYPE:"product-review")'

@@

We used R v4.1.2 [@R-base] for searches, data handling, analysis and reporting. The searches and data export from the Europe PMC were conducted with the europepmc package [@europepmc]. Indicators of transparent practices from the available full-texts were extracted with the rtransparent package [@rtransparent]. Trends over time in transparent practices by journal- and article-related information were reported using descriptive tabulations and graphical illustrations, for instance using the ggplot2 package [@ggplot2]. We used the numbers for sensitivity and specificity of the rtransparent package (3) to generate 95% confidence intervals (CIs) for our prevalence estimates of the transparency practices with the epiR package [@epiR]. We also provided monthly trends for transparency practices, to check whether there were any trends in transparent practices over time. We used Chi-square for comparing adherence to practices between years and the Wilcoxon rank sum test to test whether there was a relationship of transparency practices with citations to articles and JIFs. The thorough system information for the analysis is available in Appendix 2.

@@

The total number of COVID-19-related articles (open access and non-open access) was 175 550, of which 133 470 (76.0%) were accessible as open-access in the EPMC. However, 852 (0.6%) were not downloadable because of technical issues and thus were excluded from our analyses. Of those 132618 remaining articles, 47550 were published in 2020, 75857 in 2021, and 9211 in the first two months of 2022 (Appendix 3).

These articles were published in 6084 journals of which the top five were International Journal of Environmental Research and Public Health (n=3711), PLoS ONE (n=3059), Scientific Reports (n=1717), Frontiers in Public Health (n=1351), and Frontiers in Psychology (n=1341). The mean and the median of the number of citations to these articles were 12.0 (standard deviation, SD = 92.27) and 2 (interquartile range, IQR = 7). The highest cited article with 15549 citations was an article published in the Lancet in 2020 [@huang2020].

About nine-tenth of the articles (n=115405) had a statement to disclose COI (87.0%, 95% CI: 86.8%--87.2%). Articles published in 2022 had a higher rate of COI disclosure (91.4% vs 89.1% for 2021, and 82.9% for 2020, P \< 0.001). More than two-thirds of the references had a funding statement (n=91425) (68.9%, 95% CI: 68.7%--69.2%). Articles in 2022 had a funding statement more frequently than two other years (79.0% vs 73.1% for 2021, and 60.4% for 2020, P \< 0.001). Less than 1 in 25 of the articles (n=4708) were registered beforehand (3.6%, 95% CI: 3.5%--3.7%) and the rate was higher in 2022 (5.4% vs 4.2% for 2021, and 2.2% for 2020, P \< 0.001).

Regarding data and code availability, less than 1 in 11 articles had shared their data (n=11552) (8.7%, 95% CI: 8.6%--8.9%). Articles in 2022 had shared data more frequently (11.9% vs 9.6% for 2021, and 6.7% for 2020, P \< 0.001). Less than 1 in 30 of the articles (n=4055) had shared their code (3.1%, 95% CI: 3.0%--3.2%) and the rate was higher in 2022 (4.0% vs 3.5% for 2021, and 2.2 for 2020, P \< 0.001).

A total of 38383 articles were published in journals with JIF. Papers that had a COI disclosure, funding statement and were registered tend to be published in journals with lower JIF (P = 0.003, \< 0.001, and = 0.007, respectively). Detailed information is illustrated in Table 1.

Table 1. Citations and journal impact factor by transparent practices.

|                    | Citations to article |         |          | Journal Impact Factor |            |          |
|-----------|-----------|-----------|-----------|:----------|:----------|-----------|
| Measurement        | Median (IQR)         |         | P-Value  | Median (IQR)          |            | P-value  |
|                    | With                 | Without |          | With                  | Without    |          |
| COI disclosure     | 2 (6)                | 2 (7)   | \< 0.001 | 4.1 (2.31)            | 7.2 (5.12) | \< 0.001 |
| Funding disclosure | 2 (6)                | 2 (8)   | \< 0.001 | 4.1 (2.32)            | 4.1 (2.94) | \< 0.001 |
| Registration       | 1 (5)                | 2 (6)   | \< 0.001 | 3.8 (2.38)            | 4.2 (2.44) | \< 0.001 |
| Data sharing       | 2 (8)                | 2 (6)   | \< 0.001 | 4.0 (2.98)            | 4.1 (2.36) | \< 0.001 |
| Code sharing       | 2 (8)                | 2 (6)   | 0.70     | 4.4 (4.41)            | 4.1 (2.32) | 0.08     |

: P-value based on the Wilcoxon rank sum test.\
COI: Conflict of interest; SD: Standard deviation; IQR: Inter-quartile range.

One of eleven (9%, n=11890) articles did not adhere to any of the five transparency practices, 24.0% (n=31 844) adhered to one, 55.9% (n=74085) adhered to two and 9.1% (n=12 110) adhered to three. 2.0% (n=2644) articles showed adherence to four transparent practices, and less than 0.1% of articles (n=45) had used all five practices.

Figure 1 shows the monthly trend for three transparent practices. All practices showed increasing monthly trend. However, the increasing trends for COI and funding disclosures were steeper compared to three other practices.

@@

Adherence to transparent practices was at similar level in COVID-19 related literature than in other biomedical literature analyzed with the same methods (see Fig 3 in \[@serghiou2021\]). This is surprising particularly when considering worldwide, remarkable and noble initiatives related to enhancing open science to tackle the pandemic. As early as in January 2020, over 100 organizations, including journals, publishers, funders, universities and other institutions, signed a statement that aimed to ensure free access to research data, tools and other information related to COVID-19 \[@vuong2022; @coronavi\]. Later also other initiatives to support the goal emerged, for instance COVID-19 Open Research Dataset (CORD-19)[@cord-19], a free resource of over 280,000 articles about the COVID-19 virus. However, it is possible that the algorithms used here did not efficiently detect all different ways of sharing data and material emerged after the pandemic, because the algorithms were validated before the pandemic [@serghiou2021].

In comparison to other studies on the transparency of COVID-19 related research is difficult due to different methodologies and scale of our research. However, it seems that data sharing was little less common than in our sample than what was the proportion detected in the beginning of the pandemic measured by the investigation of associated data in PubMed Central \[@lucas-dominguez2021\]. In addition, the proportion of studies which shared data in our study sample was lower than in COVID-19 related preprints shared via medRxiv and bioRxiv \[@strcic2022\]. Lacking protocol registration, sharing code and COI have got attention during the pandemic \[@Glasziou2020; @besançon2021; @luengo-oroz2020\], however we are not aware of the investigation of transparency in these aspects in COVID-19 related research in this scale in medical research.

@@

>>>>

format:
  docx:
    section-numbers: true
    highlight-style: github 
#format: html
#format:
#  pdf:
#    colorlinks: true
editor: visual
from: markdown+tex_math_single_backslash
bibliography: references.bib
csl: vancouver.csl
---

@@

**Results:** The total number of COVID-19-related articles was 175,550, of which we were able to retrieve full texts of 132,598 (76%) articles from the database Of those, 87.0%, (95% CI: 86.8%--87.2%) provided conflicts of interest disclosure and 68.9% (95% CI: 68.7%--69.2%) funding disclosure and 3.6%, (95% CI: 3.5%--3.7%) had registered protocol. 8.7% (95% CI: 8.6%--8.9%) of studies shared raw data and 3.1% (95% CI: 3.0%--3.2%) of studies shared code. Transparent practices were more common in articles published in 2022 than in 2021 and 2020. Apart from COI disclosure, articles that adhered to transparency practices were published in journals with slightly higher median Journal Impact Factor but received equal citations.

@@

During the recent COVID-19 pandemic, part of the public in developed countries justified their lack of adherence to public health measures due to the lack of transparency of the scientific studies used as an argument by national authorities [@Skipper2020-rs]. Thus, supranational bodies such as the OECD indicate that transparent communication to citizens is vital to support public health measures and to counter misinformation [@Bellantoni_undated-tf]. On the other hand, the COVID-19 pandemic launched an incredibly fast and voluminous avalanche of scientific publications across the different fields of science. This research led to the development of vaccines as well as the evaluation of community interventions. On the one hand, it is possible that transparent scientific practices, like data and code sharing, have helped to fight pandemic globally [@besançon2021; @vuong2022], while on the other hand, the lack of transparency of some key developments has led the public to mistrust the results of research and public health measures [@Skipper2020-rs]. This shows that there is a lack of awareness about the level of transparency practices in COVID-19-related medical literature. As an example, it is unknown whether open science initiatives related to COVID-19 pandemic [@besançon2021; @vuong2022] to advance data sharing, has led to higher number of articles sharing data in COVID-19 research compared to what has been seen in studies in general in biomedical literature [@serghiou2021].

@@

We searched for open access COVID-19-related PubMed-indexed articles available in the EPMC database. We restricted our search to papers in English and considered papers published from 2020-01-01 to 2021-12-31 with variants of COVID-19 keyword in their title, keywords, or results. This enabled us to bypass screening for COVID-19-related papers and decrease the human labor for our research. We confirmed our approach by choosing a random sample of 500 papers and of those, all but 10 were relevant to COVID-19.

@@

(COVID-19 query for title, keywords, and results) AND (SRC:"MED") AND (LANG:"eng" OR LANG:"en" OR LANG:"us") AND (FIRST_PDATE:[2020-01-01 TO 2021-12-31]) AND (OPEN_ACCESS:y) AND (PUB_TYPE:"Journal Article" OR PUB_TYPE:"research-article" OR PUB_TYPE:"rapid-communication" OR PUB_TYPE:"product-review")'

@@

We used R v4.1.2 [@base] for searches, data handling, analysis and reporting. The searches and data export from the Europe PMC were conducted with the europepmc package [@europepmc]. Indicators of transparency practices from the available full-texts were extracted with the rtransparent package [@rtransparent]. Trends over time in transparency practices were reported using descriptive tabulations and graphical illustrations using the ggplot2 package [@ggplot2]. We used the sensitivity and specificity of the rtransparent package [@serghiou2021] to generate 95% confidence intervals (CIs) for our prevalence estimates of the transparency practices with the epiR package [@epiR].

Additionally, we calculated how many transparency practices articles had adhered to (from 0 to 5 practices). We used Chi-square for comparing indicators between years and the Wilcoxon rank sum test to test whether there was a relationship between transparency indicators and JIF or received citations. Fisher exact tests with Monte-Carlo simulated p-values for differences in transparency practices by journals and publishers were performed. The thorough information, data and code for the analysis are available as supplementary material on the OSF repository ([osf.io/x3kb6](https://osf.io/x3kb6/)).

@@

The total number of COVID-19-related articles (open access and non-open access) was 175,544, of which 133,467 (76.0%) were accessible as open-access in the EPMC. However, 869 (0.7%) were not downloadable because of technical issues and thus were excluded from our analyses. Of those 132,598 remaining articles, 47,548 were published in 2020, 75,842 in 2021, and 9,208 in the first two months of 2022.

These articles were published in 6,020 journals of which the top five were International Journal of Environmental Research and Public Health (n=3,710), PLoS ONE (n=3,059), Scientific Reports (n=1,716), Frontiers in Public Health (n=1,351), and Frontiers in Psychology (n=1,341). The mean and the median of the number of citations to these articles were 12.0 (standard deviation, SD = 92.28) and 2 (interquartile range, IQR = 7). The highest cited article with 15,549 citations was an article published in the Lancet in 2020 [@huang2020].

About nine-tenth of the articles (n=115,387) had a statement to disclose COI (87.0%, 95% CI: 86.8%--87.2%). Articles published in 2022 had a higher rate of COI disclosure (91.4% (2022), 89.1% (2021), 82.9% (2020), P\<0.001). More than two-thirds of the references had a funding statement (n=91,413) (68.9%, 95% CI: 68.7%--69.2%). Articles in 2022 had a funding statement more frequently than two other years (79.0% (2022), 73.1% (2021), 60.4% (2020), P\<0.001). Less than 1 in 25 of the articles (n=4708) were registered beforehand (3.6%, 95% CI: 3.5%--3.7%) and the rate was higher in 2022 (5.4% (2022), 4.2% (2021), 2.2% (2020), P\<0.001).

Regarding data and code availability, less than 1 in 11 articles had shared their data (n=11551) (8.7%, 95% CI: 8.6%--8.9%). Articles in 2022 had shared data more frequently (11.9% (2022), 9.6% (2021), 6.7% (2020), P\<0.001). Less than 1 in 30 of the articles (n=4055) had shared their code (3.1%, 95% CI: 3.0%--3.2%) and the rate was higher in 2022 (4.0% (2022), 3.5% (2021), 2.2 (2020), P\<0.001).

There was a consistently increasing monthly trend in the proportion of articles with COI disclosure, funding disclosure and protocol registration, and the proportions were the highest in February 2022 (Figure 1B). Increases in data and code sharing were weaker than in other transparency practices over time.

One-eleventh of the articles (9.0%, n=11,889) did not adhere to any of the five transparency practices, 24.0% (n=31,836) adhered to one, 56.0% (n=74,075) adhered to two, and 9.1% (n=12,109) adhered to three. Few (2.0%, n=2,644) articles showed adherence to four transparency practices, and less than 0.1% of the articles (n=45) had used all five practices.

There were journal and publisher-related statistically significant differences in all five practices (p\<0.001 for all), except for code sharing based on JIF (0.14) (Table 1 and 2, Supplementary material).

Apart from COI disclosure, articles that adhered to transparency practices were published in journals with slightly higher median JIF but received equal citations (Table 3).

@@

Adherence to transparent practices was at similar level in COVID-19 related literature than in other biomedical literature analyzed with the same methods (see Fig 3 in [@serghiou2021]). This is surprising particularly when considering worldwide, remarkable and noble initiatives related to enhancing open science to tackle the pandemic. As early as in January 2020, over 100 organizations, including journals, publishers, funders, universities and other institutions, signed a statement that aimed to ensure free access to research data, tools and other information related to COVID-19 [@vuong2022; @coronavi]. Later also other initiatives to support the goal emerged, for instance COVID-19 Open Research Dataset (CORD-19)[@cord-19], a free resource of over 280,000 articles about the COVID-19 virus. However, it is possible that the algorithms used here did not efficiently detect all different ways of sharing data and material emerged after the pandemic, because the algorithms were validated before the pandemic [@serghiou2021].

In comparison to other studies on the transparency of COVID-19 related research is difficult due to different methodologies and scale of our research. However, it seems that data sharing was little less common than in our sample than what was the proportion detected in the beginning of the pandemic measured by the investigation of associated data in PubMed Central [@lucas-dominguez2021]. In addition, the proportion of studies which shared data in our study sample was lower than in COVID-19 related preprints shared via medRxiv and bioRxiv [@strcic2022]. Lacking protocol registration, sharing code and COI have got attention during the pandemic [@Glasziou2020; @besançon2021; @luengo-oroz2020], however we are not aware of the investigation of transparency in these aspects in COVID-19 related research in this scale in medical research.

@@

# Tables

Table 1. Transparency practices in the six most common journals in the sample.

| **Measurement**       | **BMC Oral Health, N = 2503** | **Dental Press J Orthod, N = 530** | **Head Face Med, N = 572** | **J Appl Oral Sci, N = 1440** | **J Indian Prosthodont Soc, N = 495** | **Med Oral Patol Oral Cir Bucal, N = 1174** | **Other, N = 3945** | **p-value** |
|--------|--------|--------|--------|--------|--------|--------|--------|--------|
| COI disclosure        | 2489 (99%)                    | 502 (95%)                          | 537 (94%)                  | 303 (21%)                     | 301 (61%)                             | 869 (74%)                                   | 3172 (80%)          | \<0.001     |
| Funding disclosure    | 2232 (89%)                    | 87 (16%)                           | 280 (49%)                  | 656 (46%)                     | 301 (61%)                             | 497 (42%)                                   | 2505 (63%)          | \<0.001     |
| Protocol registration | 326 (13%)                     | 12 (2.3%)                          | 23 (4.0%)                  | 31 (2.2%)                     | 4 (0.8%)                              | 43 (3.7%)                                   | 298 (7.6%)          | \<0.001     |
| Data sharing          | 120 (4.8%)                    | 2 (0.4%)                           | 2 (0.3%)                   | 7 (0.5%)                      | 0 (0%)                                | 5 (0.4%)                                    | 76 (1.9%)           | \<0.001     |
| Code sharing          | 6 (0.2%)                      | 0 (0%)                             | 0 (0%)                     | 0 (0%)                        | 0 (0%)                                | 0 (0%)                                      | 0 (0%)              | 0.61        |

P-values from Fisher's Exact Test for Count Data with simulated p-value (based on 2000 replicates). COI: Conflict of interest.

Table 2. Transparency practices in the six most common publishers of journals in the sample.

| Measurement           | BioMed Central, N = 3075 | Dental Press Editora, N = 530 | Faculdade de Odontologia de Bauru, N = 1440 | Medicina Oral, Patologia Oral y Cirugia Bucal, N 1174 | Springer Verlag, N = 935 | Wolters Kluwer Medknow Publications, N = 495 | Other, N = 2619 | p-value |
|--------|--------|--------|--------|--------|--------|--------|--------|--------|
| COI disclosure        | 3026 (98%)               | 502 (95%)                     | 303 (21%)                                   | 869 (74%)                                             | 904 (97%)                | 301 (61%)                                    | 1878 (72%)      | \<0.001 |
| Funding disclosure    | 2512 (82%)               | 87 (16%)                      | 656 (46%)                                   | 497 (42%)                                             | 596 (64%)                | 301 (61%)                                    | 1593 (61%)      | \<0.001 |
| Protocol registration | 349 (11%)                | 12 (2.3%)                     | 31 (2.2%)                                   | 43 (3.7%)                                             | 113 (12%)                | 4 (0.8%)                                     | 144 (5.5%)      | \<0.001 |
| Data sharing          | 122 (4.0%)               | 2 (0.4%)                      | 7 (0.5%)                                    | 5 (0.4%)                                              | 13 (1.4%)                | 0 (0%)                                       | 54 (2.1%)       | \<0.001 |
| Code sharing          | 6 (0.2%)                 | 0 (0%)                        | 0 (0%)                                      | 0 (0%)                                                | 1 (0.1%)                 | 0 (0%)                                       | 1 (\<0.1%)      | 0.68    |

P-values from Fisher's Exact Test for Count Data with simulated p-value (based on 2000 replicates). 391 articles were published in journal without available publisher information. COI: Conflict of interest.

Table 3. Citations to article and journal impact factor by transparency practices.

|                    | Citations to article |         |          | Journal Impact Factor |            |          |
|-----------|-----------|-----------|-----------|:----------|:----------|-----------|
| Measurement        | Median (IQR)         |         | P-Value  | Median (IQR)          |            | P-value  |
|                    | With                 | Without |          | With                  | Without    |          |
| COI disclosure     | 2 (5)                | 5 (9)   | \< 0.001 | 2.8 (0.61)            | 2.7 (0.06) | \< 0.001 |
| Funding disclosure | 2 (5)                | 3 (7)   | \< 0.001 | 2.8 (0.06)            | 2.7 (0.70) | \< 0.001 |
| Registration       | 1 (4)                | 2 (7)   | \< 0.001 | 2.8 (0.01)            | 2.7 (0.61) | \< 0.001 |
| Data sharing       | 2 (4)                | 2 (7)   | 0.01     | 2.8 (0.00)            | 2.8 (0.49) | \< 0.001 |
| Code sharing       | 1.5 (3)              | 2 (7)   | 0.70     | 2.8 (0.00)            | 2.8 (0.49) | 0.08     |

P-value based on the Wilcoxon rank sum test. 1929 articles were published in journals with no impact factor. COI: Conflict of interest; SD: Standard deviation; IQR: Inter-quartile range.
